### 2025-08-07 actuarial.qmd
20250807 - additions and tweaks

As an example of using the actuarial approach, Billy Beane, who was the general manager of the Oakland Athletics at the time, wanted to find ways for his team—which had less finanical resources than its competitors—to compete with teams that hard more money to sign players.
Because the team did not have the resources to sign the best players, they had to find to find other ways to find the optimal players that they could afford.
So, he used statistical formulas that weight variables, such as on-base percentage and slugging percentage, according to their value for winning games, for the use of selecting players.
His approach became well-known based on the Michael Lewis book, "Moneyball: The Art of Winning an Unfair Game", and the eventual movie, "Moneyball".

@@

There are numerous ways in which humans and statistical algorithms could be involved.
On one extreme, humans make all judgments.
On the other extreme, although humans may be involved in data collection, a statistical formula makes all decisions based on the input data, consistent with an actuarial approach.
However, the human judgment and actuarial approaches can be combined in a hybrid way [@Dana2006].
For example, to save time and money, a clinical psychologist might use an actuarial approach in all cases, but might only use a judgment approach when the actuarial approach gives a "positive" test.
Or, the clinical psychologist might use both human judgment and an actuarial approach independently to see whether they agree.
That is, the clinician may make a prediction based on their judgment and might also generate a prediction from an actuarial approach.

@@

There *may* be situations in which a human judgment would do better than an actuarial judgment.
One situation where human judgment would be important is when no actuarial method exists for the judgment or prediction.
For instance, when no actuarial method exists for the diagnosis or disorder (e.g., suicide), it is up to the clinician.
However, we could collect data on the outcomes or on clinicians' judgments to develop an actuarial method that will be more reliable than the clinicians' judgments.
That is, an actuarial method developed based on clinicians' judgments will be more accurate than clinicians' judgments.
In other words, we do not necessarily need outcome data to develop an actuarial method.
We could use the client's data as predictors of the clinicians' judgments to develop a structured approach to prediction that weighs factors similarly to clinicians, but with more [reliable](#sec-reliability) predictions.

@@

Another situation in which human judgment could be preferable is if advanced, complex theories exist.
Computers have a difficult time adhering to complex theories, so clinicians may be better suited.
However, we do not have any of these complex theories in psychology that are accurate.
We would need strong theory informed by data regarding causal influences, and [accurate](#validity) measures to assess them.
However, no theories in psychology are that good.
Nevertheless, predictive accuracy can be improved when considering theory [@Silver2012; @Garb2019].

If the prediction requires complex configural relations that a computer will have a difficult time replicating, a clinician's judgment may be preferred.
Although the likelihood that a person can accurately work through these complex relations is theoretically possible, it is highly unlikely.
Holistic pattern recognition (such as language and faces) tends to be better by humans than computers.
But computers are getting better with holistic pattern recognition through machine learning.

@@

Hundreds of studies have examined clinical versus actuarial prediction methods across many disciplines.
Findings consistently show that actuarial methods are as [accurate](#sec-validity) or more [accurate](#sec-validity) than human judgment/prediction methods.
"There is no controversy in social science that shows such a large body of qualitatively diverse studies coming out so uniformly...as this one" [@Meehl1986, pp. 373–374].

@@

Despite actuarial methods being generally more accurate than human judgment, judgment is much more widely used by clinicians.
There are several reasons why actuarial methods have not caught on; one reason is professional traditions.
Experts in any field do not like to think that a computer could outperform them.
Some practitioners argue that judgment/prediction is an "art form" and that using a statistical formula is treating people like a number.
However, using an approach (i.e., human judgment) that systematically leads to less [accurate](#sec-validity) decisions and predictions is an ethical problem.

Some clinicians do not think that group averages (e.g., in terms of which treatment is most effective) apply to an individual client.
This invokes the distinction between nomothetic (group-level) inferences and idiographic (individual-level) inferences.
However, the scientific evidence and probability theory strongly indicate that it is better to generalize from group-level evidence than throwing out all the evidence and taking the approach of "anything goes."
Clinicians frequently believe the broken leg fallacy, i.e., thinking that your client is an exception to the algorithmic prediction.
In most cases, deviating from the statistical formula will result in less [accurate](#sec-validity) predictions.
People tend to overestimate the probability of low [base rate](#sec-baseRates) conditions and events.

@@

Another argument against using actuarial methods is that "no methods exist".
In some cases, that is true—actuarial methods do not yet exist for some prediction problems.
However, one can always create an algorithm of the experts' judgments, even if one does not have access to the outcome information.
A model of clinicians' responses tends to be more [accurate](#sec-validity) than clinicians' judgments themselves because the model gives the same outcome with the same input data—i.e., it is perfectly [reliable](#sec-reliability).

@@

>>>>

As an example of using the actuarial approach, Billy Beane, who was the general manager of the Oakland Athletics at the time, wanted to find ways for his team—which had less financial resources than its competitors—to compete with teams that had more money to sign players.
Because the team did not have the resources to sign the best players, they had to find to find other ways to find the optimal players that they could afford.
So, he used statistical formulas that weight variables, such as on-base percentage and slugging percentage, according to their value for winning games, for the use of selecting players.
His approach became well-known based on the Michael Lewis book, "Moneyball: The Art of Winning an Unfair Game", and the eventual movie, "Moneyball".

@@

There are numerous ways in which humans and statistical algorithms could be involved.
On one extreme, humans make all judgments.
On the other extreme, although humans may be involved in data collection, a statistical formula makes all decisions based on the input data, consistent with an actuarial approach.
However, the human judgment and actuarial approaches can be combined in a hybrid way [@Dana2006].
For example, to save time and money, a clinical psychologist might use an actuarial approach in all cases, but might only use a judgment approach when the actuarial approach gives a "positive" test.
Or, the clinical psychologist might use both human judgment and an actuarial approach independently to see whether they agree.
That is, the clinician or expert may make a prediction based on their judgment and might also generate a prediction from an actuarial approach.

@@

There *may* be situations in which a human judgment would do better than an actuarial judgment.
One situation where human judgment would be important is when no actuarial method exists for the judgment or prediction.
For instance, when no actuarial method exists for the diagnosis or disorder (e.g., suicide), it is up to the clinician/expert.
However, we could collect data on the outcomes or on experts' judgments to develop an actuarial method that will be more reliable than the experts' judgments.
That is, an actuarial method developed based on experts' judgments will be more accurate than experts' judgments.
In other words, we do not necessarily need outcome data to develop an actuarial method.
We could use the player's data as predictors of the experts' judgments to develop a structured approach to prediction that weighs factors similarly to experts, but with more [reliable](#sec-reliability) predictions.

@@

Another situation in which human judgment could be preferable is if advanced, complex theories exist.
Computers have a difficult time adhering to complex theories, so clinicians/experts may be better suited.
However, we do not have any of these complex theories in psychology that are accurate.
We would need strong theory informed by data regarding causal influences, and [accurate](#validity) measures to assess them.
However, no theories in psychology are that good.
Nevertheless, predictive accuracy can be improved when considering theory [@Silver2012; @Garb2019].

If the prediction requires complex configural relations that a computer will have a difficult time replicating, an expert's judgment may be preferred.
Although the likelihood that a person can accurately work through these complex relations is theoretically possible, it is highly unlikely.
Holistic pattern recognition (such as language and faces) tends to be better by humans than computers.
But computers are getting better with holistic pattern recognition through machine learning.

@@

Hundreds of studies have examined clinical versus actuarial prediction methods across many disciplines.
Despite the romantic notion that professional scouts and experts being able to make accurate predictions (and identify strong players), findings consistently show that actuarial methods are as [accurate](#sec-validity) or more [accurate](#sec-validity) than human judgment/prediction methods [@Dawes1989; @Aegisdottir2006; @Baird2000; @Grove1996; @Grove2000].
"There is no controversy in social science that shows such a large body of qualitatively diverse studies coming out so uniformly...as this one" [@Meehl1986, pp. 373–374].
In general, actuarial methods tend to be about 10% more accurate than human judgment [@Grove2000].
Thus, actuarial methods have [incremental validity](#sec-incrementalValidity) over human judgment.
As noted earlier, the superiority of actuarial approaches to clinical judgment tends to hold even when the expert is given more information than the actuarial approach [@Dawes1989].
In addition, actuarial predictions outperform human judgment even when the human is given the result of the actuarial prediction [@Kahneman2011].
Allowing clinicians/experts to override actuarial predictions consistently leads to lower predictive accuracy [@Garb2019].

@@

Despite actuarial methods being generally more accurate than human judgment, judgment is much more widely used by clinicians/experts.
There are several reasons why actuarial methods have not caught on; one reason is professional traditions.
Experts in any field do not like to think that a computer could outperform them.
Some practitioners argue that judgment/prediction is an "art form" and that using a statistical formula is treating people like a number.
However, using an approach (i.e., human judgment) that systematically leads to less [accurate](#sec-validity) decisions and predictions is an ethical problem.

Some experts do not think that group averages (e.g., in terms of which treatment is most effective) apply to an individual client or player.
This invokes the distinction between nomothetic (group-level) inferences and idiographic (individual-level) inferences.
However, the scientific evidence and probability theory strongly indicate that it is better to generalize from group-level evidence than throwing out all the evidence and taking the approach of "anything goes."
Clinicians frequently believe the broken leg fallacy, i.e., thinking that your client is an exception to the algorithmic prediction.
In most cases, deviating from the statistical formula will result in less [accurate](#sec-validity) predictions.
People tend to overestimate the probability of low [base rate](#sec-baseRates) conditions and events.

@@

Another argument against using actuarial methods is that "no methods exist".
In some cases, that is true—actuarial methods do not yet exist for some prediction problems.
However, one can always create an algorithm of the experts' judgments, even if one does not have access to the outcome information.
A model of experts' responses tends to be more [accurate](#sec-validity) than experts' judgments themselves because the model gives the same outcome with the same input data—i.e., it is perfectly [reliable](#sec-reliability).

@@

Another potential reason clinicians may not use actuarial judgment is time.
This can be a legitimate challenge to using actuarial models in some contexts.
Actuarial methods can take time to apply—they require assessing the releving criteria, putting the assessment scores into an actuarial formula, and interpreting the output score.
In some clinical situations, such as risk assessments in terms of whether someone is imminently suicidal or homicidal and how to proceed, clinicians must make immediate in-session decisions @Jacinto2018.
Thus, it can be challenging or impractical in some situations to conduct all of the relevant assessments using actuarial approaches.
Moreover, in some workplace settings, such as health care settings, clinicians may be pushed beyond their limits in terms of productivity expectations.
In these situations, one possible approach is to exercise their best possible judgment based on the available evidence [i.e., intuitive decision-making processes; @Jacinto2018], but to check it against an actuarial model after the fact.
It is valuable to evaluate to receive feedback on the accuracy of one's judgments and, when time permits, to evaluate one's intuitive decision-making against analytical decision-making.
